Global code motion of congruent instructions

### 2017-11-06 cgo/paper.tex
sinking

There has been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using integer linear programming (ILP) which results in prohibitively
high compile time which is not suitable for industrial compilers. To the best of
our knowledge we have not found any reference which explored \gcm{} of congruent
instructions. A part of our implementation (aggressive code hoisting) is already
merged in \LLVM{} trunk. The main contributions of this paper are:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item an optimistic fast algorithm to schedule congruent instructions;
\item a cost model to reduce interference and hence, reduce spills;
\item experimental evaluation of our implementation in \LLVM{}.
\end{itemize}

@@

In order to be able to move an instruction that has side effects, we must
guarantee that the side-effects of that instruction before and after its
code-motion will produce the same side-effects.  For hoisting in particular, we
need to prove that on all execution paths from the hoisting point to the end of
the program the side effects appear in the same order. In general, we cannot
introduce a new computation along any path of execution without checking for
undefined behaviors \cite{undef}. Since \GCM{} is only dealing with VBEs, as
hoisting happens for CHIs with complete entries, the downward safety is ensured
by construction of CHIs.

@@

@@

\begin{verbatim}
int doCodeSinking()
  // Find sinkable instructions
  For each value number VN with 2 or more instructions:
    if there are two instructions with same immediate
        successor(S) as post-dominator:
      if both are only used in the same PHI-node of S
          && dependencies of both are available at S:
         mark instructions as sinkable

@@

Code hoisting opens new opportunities for other hoistable candidates which were
of higher rank (depended on candidates which got hoisted). Ideally we could
iterate on lower ranking expressions first and then proceed to higher ranking
expressions in the same iteration but \LLVM{'}s \GVN{} infrastructure does not
compute equivalence classes in an effective way. We found it simpler to just
recompute the value numbers and start finding hoistable candidates again.

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that are
candidates for \gcm{}, matching the complexity of \PRE{} on \SSA{} form.
The analyses computed for this pass are Global Value Numbering, Computation of
DJ-graph and merge-sets, Marking Barriers, all are linear in number of
instructions in a function. Liveness analysis is expensive and only performed
on-demand for hoistable candidates so it does not affect the overall compilation
time by much. Other analyses like Alias Analysis, \MemorySSA{} and Dominator Tree
are already available in the \LLVM{} pass pipeline. Although we recompute \GVN{}, and
Barriers for each iteration of the \gcm{}, we have not seen significant
increase in the compilation times (see Section~\ref{tab:compile-time}). We have also
provided appropriate compiler
flags to expedite \gcm{} by bailing out with fewer iterations, or skip the
liveness based profitability analysis to aggressively move the code as long as
they are legal.

The analysis is followed by a simple code generation that adds the identified
instruction in the destination point and removes all instructions rendered
redundant as a result of \gcm{}.

>>>>

There has been a lot of work both in industry and academia to hoist and sink
code out of branches, and in general global scheduling
\cite{click1995global}. Some relate code-hoisting to code-size optimization
\cite{rosen1988global} and many \cite{barany2013, shobaki2013} use global
scheduling to improve performance. Most of the recent work on global scheduling
are done using integer linear programming (ILP) which results in prohibitively
high compile time which is not suitable for industrial compilers. To the best of
our knowledge we have not found any reference which explored \gcm{} of congruent
instructions. A part of our implementation (aggressive code hoisting) is already
merged in \LLVM{} trunk. The main contributions of this paper are:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item an optimistic fast algorithm to schedule congruent instructions;
\item introducing an analogue of PHI nodes on the inverse CFG which allows
  faster tracking of anticipable instructions;
\item a cost model to reduce interference and hence, reduce spills;
\item experimental evaluation of our implementation in \LLVM{}.
\end{itemize}

@@

In order to be able to move an instruction that has side effects, we must
guarantee that the side-effects of that instruction before and after its
code-motion will produce the same side-effects.  For hoisting in particular, we
need to prove that on all execution paths from the hoisting point to the end of
the program the side effects appear in the same order. In general, we cannot
introduce a new computation along any path of execution without checking for
undefined behavior \cite{undef}. Since \GCM{} is only dealing with VBEs, as
hoisting happens for CHIs with complete entries, the downward safety is ensured
by construction of CHIs.

@@

Code hoisting opens new opportunities for other hoistable candidates which were
of higher rank (depended on candidates which got hoisted) which are encountered
subsequently as the algorithm visits instructions in the increasing order of
their ranks.

@@

For sinking we reuse the PHI nodes instead of inserting another expression nodes
like in \cite{ssapre} because PHI nodes ensure availability of definitions, and
hence downward safety, as long as all instructions corresponding to a PHI are
congruent. Also, the PHI nodes represent a complete set of instructions which
could be potentially sunk for code-size optimizations because PHI nodes
represent all values flowing out from a basic block into its dominance-frontier.

\begin{verbatim}
int doCodeSinking()
  // Find sinkable instructions
  For each value number VN > 1 instructions:
    if there are two instructions with same immediate
        successor(S) as post-dominator:
      if both are only used in the same PHI-node of S
          && dependencies of both are available at S:
         mark instructions as sinkable

@@

\subsection{Time complexity of algorithm}
The complexity of code hoisting is linear in number of instructions that are
candidates for \gcm{}, matching the complexity of \PRE{} on \SSA{} form.  The
analyses computed for this pass like Global Value Numbering, Marking Barriers,
are both linear in number of instructions in a function. Computation of DJ-graph
and merge-sets is $O(E)$, E being number of edges in the CFG. Liveness analysis
is not very expensive \cite{das2012} but still only performed on-demand for
hoistable candidates. Other analyses like Alias Analysis, \MemorySSA{} and
Dominator Tree are already available in the \LLVM{} pass pipeline.

Although we recompute \GVN{}, and Barriers for each iteration of the \gcm{},
there is still gain in the compilation times (see
Section~\ref{tab:compile-time}). We have also provided appropriate compiler
flags to expedite \gcm{} by bailing out with fewer iterations, or skip the
liveness based profitability analysis to aggressively move the code as long as
they are legal.
